---
title: "3. Advanced Mapping"
author: "Paolo Di Lorenzo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Advanced Mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will explore some of the more advanced mapping features of `usmap`. Before continuing, be sure to check out [Mapping the US](mapping.html) as that will cover more of the basics of plotting US maps and styling them with `ggplot2`.

## Labels

As of `usmap 0.4.0`, maps with state labels can be created:

```{r, fig.align='center', fig.width=7, fig.height=5, message=FALSE, warning=FALSE}
usmap::plot_usmap("states", labels = TRUE)
```

`usmap 0.5.0` adds the ability to add county labels:

```{r, fig.align='center', fig.width=7, fig.height=5, message=FALSE, warning=FALSE}
usmap::plot_usmap("counties", include = c("MA", "CT", "RI"), labels = TRUE)
```

Labels can be colored using the `label_color` parameter:

```{r, fig.align='center', fig.width=7, fig.height=5, message=FALSE, warning=FALSE}
usmap::plot_usmap("counties",
                  include = c("MA", "CT", "RI"),
                  labels = TRUE, label_color = "blue")
```

## `ggplot2` aesthetic mapping parameters

Parameters used by the map's aesthetic mapping (`ggplot2::aes`) can be passed directly via `plot_usmap` by adding the parameters anywhere at the call site:

```{r, fig.align='center', fig.width=7, fig.height=5, message=FALSE, warning=FALSE}
usmap::plot_usmap("counties",
                  include = c("MA", "CT", "RI"),
                  labels = TRUE, label_color = "blue",
                  fill = "yellow", alpha = 0.25, color = "orange", linewidth = 2)
```

Notice in this case we set the `fill` and `alpha` parameters to fill in the counties with a semi-transparent yellow color.

The following parameters are supported:

-   `fill`: fill color of the state/county polygons
-   `alpha`: transparency of the state/county polygon fill colors
-   `color`/`colour`: line color of the state/county polygons
-   `linewidth`: thickness of the state/county polygon lines
