---
title: "4. Example Analysis"
author: "Yingying Yu"
date: "2023-12-03"
output:
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Diving into US Tornado Data

This page is created for an exercise for the BSPH course Statistical Programming Paradigms and Workflows (140.777), and it demonstrates some of the functions in the package with example data from TidyTuesday. Codes are revisited from Evan Gower's post on [Kaggle](https://www.kaggle.com/code/evangower/diving-into-us-tornado-data/notebook).

In this analysis, I am trying to understand how many tornadoes happened through the years from 1950 to 2021? Where do tornadoes usually occur? What are the fatalities caused by tornadoes by states? The source data comes from NOAA's National Weather Service Storm Prediction Center [Severe Weather Maps, Graphics, and Data Page](https://www.spc.noaa.gov/wcm/#data), The data dictionary can be found [here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-05-16/readme.md).

### 0. Load the data

```{r, warning=FALSE}
if (!requireNamespace("here", quietly = TRUE)) {
  install.packages("here")
}
library(here)
library(tidyverse)
if (!file.exists(here("data", "tuesdata_drugs.RDS"))) {
    tuesdata <- tidytuesdayR::tt_load("2023-05-16")
    tornados <- tuesdata

    # save the files to RDS objects
    save(tuesdata$tornados, file = here("data", "tuesdata_tornados.rda"))
}

tornados <- load(here("data", "tuesdata_tornados.rda"))
glimpse(tornados)
```

### 1. Prepare and wrangle data

```{r}
# delete several columns
data <- tornados[, c("om","yr","mo","date","st","fat","mag")]
```

### 2. Tornadoes through the years

Before we start in are main dive, lets look to see how many tornadoes were there from the time period 1950 to 2021.

```{r}
# Count the number of tornadoes
data %>% count() %>% print()
```

```{r}
# Summarise average tornadoes by year
yearly_tornado <- data %>%
  group_by(yr) %>%
  count()

# Add custom font
library(showtext)
font_add_google("Ubuntu")
font_labels <- "Ubuntu"
showtext_auto()

# Plot data
ggplot(yearly_tornado, aes(yr, n)) +
  geom_line(color = "#98103E") +
  labs(title = "Tornado through the Years",
       subtitle = "The number of tornadoes has risen over the past seventy years",
       caption = "Period 1950 - 2021 | Data: NOAA",
       x = "",
       y = "Number of tornadoes") +
  theme_minimal() +
  theme(text = element_text(family = font_labels, color = 'black'),
        plot.title = element_text(size = 20, face = "bold"),
        plot.subtitle = element_text(size = 12),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank())
```

We can see that there is a gradually increasing trend of number of tornadoes as years increase. Is this trend similar in the top 4 cities having most tornadoes?


### 3. Where do Tornadoes Occur?

Does tornadoes more likely to occur at some states?

```{r}
# Summarise number of tornadoes by state
tornadoes_st <- data %>%
  group_by(st) %>%
  summarise(value = n()) %>%
  arrange(desc(value))

# Print result
print(tornadoes_st)
```

```{r}
yearly_tornado_top <- data %>%
  group_by(yr, st) %>%
  filter(st == "TX" | st == "KS" | st == "OK" | st == "FL") %>% 
  count()

# Plot data
ggplot(yearly_tornado_top, aes(yr, n)) +
  geom_line(color = "#98103E") +
  facet_wrap(~ st, nrow = 2) +
  labs(title = "Tornado of top cities through the Years",
       subtitle = "Pattern of the top 4 states are different",
       caption = "Period 1950 - 2021 | Data: NOAA",
       x = "",
       y = "Number of tornadoes") +
  theme_minimal() +
  theme(text = element_text(family = font_labels, color = 'black'),
        plot.title = element_text(size = 20, face = "bold"),
        plot.subtitle = element_text(size = 12),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank())
```

Texas by a clear margin saw the most tornadoes in this timeframe. Kansas, Oklahoma, and Florida round out the top four. This package also provide a function to retrieve the unique FIPS (Federal Information Processing Standards) code for a state or county. Let's try it with the top 4 states.

```{r}
library(usmap)
fips(c("Tx", "KS", "OK", "FL"))
```


```{r}
# Add color levels to data for plot
tornadoes_st <- tornadoes_st %>%
  mutate(color = case_when(value < 500 ~ "Less than 500",
                           value <= 1000 ~ "501 - 1,000",
                           value <= 2000 ~ "1,001 - 2,000",
                           value <= 4000 ~ "2,001 - 4,000",
                           value > 4000 ~ "More than 4,000",
                           TRUE ~ "No Tornadoes"))

# Arrange color levels
tornadoes_st$color <- fct_relevel(tornadoes_st$color, c("More than 4,000",
                                                        "2,001 - 4,000",
                                                        "1,001 - 2,000",
                                                        "501 - 1,000",
                                                        "Less than 500"))
# Change column name to fit the usmap setting
names(tornadoes_st)[1]<-"state"

# Plot data
plot_usmap(data = tornadoes_st, values = "color", labels = FALSE) +
  scale_fill_manual(values = c( "Less than 500" = "#ffdfa4",
                                "501 - 1,000" = "#FFC457",
                                "1,001 - 2,000"= "#E4683F",
                                "2,001 - 4,000"= "#C03434",
                                "More than 4,000" = "#98103E",
                                "No Tornadoes" = "grey70")) +
  labs(title = "Number of Tornadoes by State",
      subtitle = "Period 1950 - 2021 | Data: NOAA",
      fill = "color",
      x = "", y = "") +
  theme(plot.title = element_text(margin = margin(b = 20, t=15), 
                                family = font_labels, color = "#22222b",
                                face = "bold", size = 25, hjust = 0.5),
        plot.subtitle = element_text(family = font_labels,
                                    color = "#22222b", size = 15, 
                                    hjust = 0.5, vjust = 4),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text=element_text(family = font_labels, size = 8), 
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x=element_blank()) +
  guides(fill = guide_legend(
    label.position = "top", family = font_labels, color = "#808080", nrow = 1,
    keywidth = 2, keyheight = 0.5))
```

Looking at the visualization:

-   The Midwest and Southeast regions have seen a high amount tornadoes touch down.
-   While the Northeast and West regions saw low levels of tornadoes.

### 4. Tornadoes by Month

```{r}
# Summarise number or tornadoes by month
month <- data %>%
  group_by(mo) %>%
  summarise(value = n())

# Add months
mymonths <- c("Jan","Feb","Mar",
              "Apr","May","Jun",
              "Jul","Aug","Sep",
              "Oct","Nov","Dec")

# Order months
month$abv <- mymonths[ month$mo ]

# Plot data
ggplot(month, aes(abv, value)) +
  geom_col(fill = "#98103E") +
  scale_x_discrete(limits = month$abv) +
  labs(title = "Number of Tornadoes by Month",
       subtitle = "May saw almost 15,000 tornadoes, by far the most in any month",
       caption = "Period 1950 - 2021 | Data: NOAA",
       x = "Month",
       y = "Number of Tornadoes") +
  theme_minimal() +
  theme(text = element_text(family = font_labels, color = 'black'),
        plot.title = element_text(size = 20, face = "bold"),
        plot.subtitle = element_text(size = 12),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank())
```
### 5. Tornado fatality

How do different level of tornado affect the fatality?

```{r}
# Summarise average fatalities by type
level_fat <- data %>%
  filter(!mag == -9)
# make EF scale as factors
level_fat$mag <- factor(level_fat$mag, c("0","1","2","3","4","5"))

# plot
ggplot(level_fat, aes(mag, fat)) +
 geom_boxplot() +
 labs(title = "Tornado Fatalities by EF Rating",
      subtitle = "Higher the rating, higher the fatality and larger the variance",
      caption = "Period 1950 - 2021 | Data: NOAA",
      x = "EF Level", y = "Fatality") +
  theme_minimal() +
  theme(text = element_text(family = font_labels, color = 'black'),
        plot.title = element_text(size = 20, face = "bold"),
        plot.subtitle = element_text(size = 12),
        panel.grid.major.x = element_blank())
```

### 6. Average number of Tornado Fatalities by State

```{r, warning==FALSE}
library(purrr)
fat <- level_fat[, c("st","yr","fat")]
wide_fat <- fat %>%
  group_by(st,yr) %>% 
  summarise(total_fat = sum(fat))

# calculate the average number of tornado fatalities over years by state
split_fat <- split(wide_fat, wide_fat$st)
mean_fat <- split_fat %>% 
 map_dbl(.f = ~mean(.x$total_fat)) %>% 
  sort(decreasing = T)
mean_fat

# calculate the total number of tornado fatalities over years by state
total_fat <- split_fat %>% 
  map_int(.f = ~sum(.x$total_fat)) %>% 
  sort(decreasing = T)
total_fat
```

Alabama had the most fatalities, followed by Texas, Mississippi, and Oklahoma.

## Conclusion

From the analysis of the tornado dataset, we found there are 67,937 tornadoes touch down in the United States from 1950 through 2021. The Midwest and Southeast are impacted the most by tornadoes. In addition, there was 6,118 fatalities caused by tornadoes, and EF5 averaged the hisghest amount of fatalities by EF rating.

## Function summary

| Package name | Functions used                                    |
|--------------|---------------------------------------------------|
| `dplyr`      | `count` `group_by` `mutate` `arrange` `summarise` |
| `forcats`    | `fct_relevel`                                     |
| `ggplot2`    | `geom_boxplot` `geom_line` `geom_col` `facet_wrap`|


